---
title: "Chapter 10 - Analysing discrete data - the generalised linear model - Exercise solutions and Code Boxes"
author: "David Warton"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 10 - Analysing discrete data - the generalised linear model - Exercise solutions and Code Boxes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Exercise 10.1: Crabs on seaweed
*David and Alistair... observed the following presence (`+`) / absence (`-`)} patterns for crabs (across 10 replicates):*
Time           5  5  5  5  5  5  5  5  5  5 ... 10
Distance (m)   0  0  0  0  0  2  2  2  2  2 ... 10
Crabs          -  +  +  -  -  -  -  +  -  - ... +
*They would like to know if there is any evidence of a difference in crab presence patterns with Distance of Isolation. How should they analyse the data?*

They are looking for evidence of a difference at different values of `Dist` so will be wanting to do a hypothesis test. Because they have presence-absence data they should be thinking of a generalised linear model with binomial response, as in Code Box 10.1.


## Exercise 10.2: Do offshore wind farms affect fish communities?
*Lena and colleagues... [were] collecting paired data before and after wind farm construction, at 36 stations in each of three zones (`Wind Farm`, `North`, or `South`):*

Zone         Impact    Station    Abundance
----------   -------   --------   ----------
Wind Farm    Before    WF1        0
Wind Farm    After     WF1        0
South        Before    S1         5
South        After     S1         0
North        Before    N1         0
North        After     N1         0
Wind Farm    Before    WF2        1
Wind Farm    After     WF2        1
...          ...       ...        ...
North        After     N36        0
---------    -------   --------   ----------

*Lena wants to know if there is any evidence of a change in eel abundance at wind farm stations, as compared to others, following construction of the wind farm. How should she analyse the data?*

She is looking for evidence of a difference across different values of `Zone` when `Impact=After` so will be wanting to do a hypothesis test. This is a BACI design, so she would be particularly interested in the `Zone:Impact` interaction. Because she has count data with plenty of zeros, this should be analysed using a generalised linear model (assuming a negative binomial or Poisson response, as in Code Box 10.1).


## Exercise 10.3: Invertebrate response to bush regeneration}
*Anthony wants to evaluate how well invertebrate communities are re-establishing following bush regeneration efforts.*

 Treatment |    Count
 --------- |  -------
 C         |        0
 R         |        3
 R         |        1
 R         |        3
 C         |        1
 R         |        2
 R         |       12
 R         |        1
 R         |       18
 R         |        0

*He wants to know if there is any evidence that bush regeneration is working. How should he analyse the data?*

He is looking for evidence of an effect of revegetation so will be wanting to do a hypothesis test for a `Treatment` effect. Because he has count data, with zeros, this should be analysed using a generalised linear model (assuming a negative binomial or Poisson response, as in Code Box 10.1).


## Code Box 10.1: Example GLM fits for Exercises 10.1-3.}
For Exercise 10.1:
```{r box10.1.1}
library(ecostats)
data(seaweed)
seaweed$CrabPres = seaweed$Crab>0
ft_crab = glm(CrabPres~Time*Dist, family=binomial("cloglog"),
             data=seaweed)
```

For Exercise 10.2:
\begin{verbatim}
```{r box10.1.2}
data(windFarms)
eels = windFarms$abund[,16]
ft_eels = glm(eels~Station+Year*Zone,family="poisson",
             data=windFarms$X)
```

For Exercise 10.3:
```{r box10.1.3}
data(reveg)
Haplotaxida=reveg$abund[,12]
library(mvabund)
worms = reveg$abund$Haplotaxida
ft_worms = manyglm(worms~treatment,family="negative.binomial", data=reveg)
```


