---
title: "Chapter 8 -- Wiggly Models -- Exercise solutions and Code Boxes"
author: "David Warton"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Chapter 8 -- Wiggly Models -- Exercise solutions and Code Boxes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Code Box 8.1: Fitting a spline smoother to the Mauna Loa annual data of Exercise 8.1 on `R`.
```{r box81}
library(mgcv)
library(ecostats)
data(maunaloa)
maunaJan = maunaloa[maunaloa$month==1,]
ft_maunagam=gam(co2~s(year), data=maunaJan)
summary(ft_maunagam)
```

## Code Box 8.2: Residual plot from a GAM of the annual Mauna Loa data of Exercise 8.1
```{r box82}
plotenvelope(ft_maunagam)
```

*Do you think assumptions are satisfied?*

It generally looks good, although there are a couple of lines of points fanning out on the residuals vs fits plot for `co2<320`.  (The method of taking measurements actually changed in 1974, at about `co2=330`, so it is possible that this could be artefacts due to changes in the measurement or recording process for data in the 1950's and 1960's.)

Note the simulation envelope around the trend line in the residual vs fits plot is hard to see -- this is because it falls very close to the horizontal line at zero. The smoother against year was actually fitted such that the residual trend would be a horizontal line at zero.


## Code Box 8.3: Comparing curves for the Mauna Loa data.

```{r box83 bic}
maunaJan$year00 = pmax(2000,maunaJan$year)
ft_maunaPiece = lm(co2~year+year00,data=maunaJan)
ft_maunagam20 = gam(co2~s(year,k=20), data=maunaJan)
summary(ft_maunagam20)$edf # this gam added about 6 extra knots:
BIC(ft_maunaPiece,ft_maunagam,ft_maunagam20)
```

```{r box83 validation}
isTrain = which(maunaJan$year<=2006)
datTrain = maunaJan[isTrain,]
datTest = maunaJan[-isTrain,]
ft_piece = lm(co2~year+year00,dat=datTrain)
ft_gam = gam(co2~s(year),dat=datTrain)
ft_gam20 = gam(co2~s(year,k=20),dat=datTrain)
pr_piece = predict(ft_piece,newdata=datTest)
pr_gam = predict(ft_gam,newdata=datTest)
pr_gam20 = predict(ft_gam20,newdata=datTest)
preds = cbind( predict(ft_piece,newdata=datTest),
    predict(ft_gam,newdata=datTest), predict(ft_gam20,newdata=datTest) )
print( apply((datTest$co2-preds)^2,2,sum)) # getting SS by column
```

## Exercise 8.2: Eucalypt richness as a function of the environment
*How does \emph{Myrtaceae} species richness vary from one area to the next, and what are the main environmental correlates of richness? [Ian] thinks species richness could respond to temperature and rainfall in a non-linear fashion, and may interact. What sort of model should Ian consider using?*

He could try a GAM, as in Code Box 8.4. Note that this doesn't allow him to account for spatial autocorrelation at the same time, maybe he could try `gamm` as a way forward.

## Code Box 8.4: Handling interactions in a GAM, when modelling Ian's richness data as a function of minimum temperature and rainfall, for Exercise 8.2

```{r box84, fig.width=6,fig.height=6}
data(Myrtaceae)
ft_tmprain=gam(log(richness+1)~te(TMP_MIN,RAIN_ANN),data=Myrtaceae)
vis.gam(ft_tmprain,theta=-135) #rotating the plot to find a nice view
summary(ft_tmprain)$edf
```

```{r box84 quad}
ft_tmprain2=gam(log(richness+1)~s(TMP_MIN)+s(RAIN_ANN)+TMP_MIN*RAIN_ANN,
      data=Myrtaceae)
summary(ft_tmprain2)
```

## Exercise 8.3: Smoothers for climate effects on plant height
*[Angela] would like to know how plant height relates to climate -- in particular, are there interactions between predictors, and is the relationship non-linear?*

```{r ex83}
data(globalPlants)

ft_1temprain = gam(log(height)~temp+rain, dat=globalPlants) #linear model
ft_2tempPlusrain = gam(log(height)~poly(temp,2)+poly(rain,2), dat=globalPlants) #quadratic, no interaction
ft_2temprain = gam(log(height)~poly(temp,2)+poly(rain,2)+rain:temp, dat=globalPlants) #quadratic, interactions
ft_stempPlusrain = gam(log(height)~s(temp)+s(rain), dat=globalPlants) #smoother+no interaction
ft_stemprain = gam(log(height)~s(temp)+s(rain)+rain:temp, dat=globalPlants) #smoother+interaction
BIC(ft_1temprain,ft_2tempPlusrain,ft_2temprain,ft_stempPlusrain,ft_stemprain)
```

And the winner is a linear model.


## Code Box 8.5: Residual plot with a smoother to diagnose a model.
```{r box85, fig.width=5}
data(globalPlants)
globalPlants$logHt = log(globalPlants$height)
ft_heightlm = lm(logHt~lat,dat=globalPlants)
plot(ft_heightlm,which=1)
ft_temp = gam(logHt~s(temp), dat=globalPlants)
ecostats::plotenvelope(ft_temp, which=1, main="")
```

## Exercise 8.4: Nonlinear predictors of species richness?
*Refit the model using smoothers for each environmental variable and compare to a model with additive quadratic terms (assume conditional independence for the moment).*

```{r ex84 smooth}
data(Myrtaceae)
Myrtaceae$logrich=log(Myrtaceae$richness+1)
ft_richAdd = lm(logrich~soil+poly(TMP_MAX,degree=2)+
     poly(TMP_MIN,degree=2)+poly(RAIN_ANN,degree=2), data=Myrtaceae)
ft_richSmooth = gam(logrich~soil+s(TMP_MAX)+s(TMP_MIN)+s(RAIN_ANN),
                    data=Myrtaceae)
BIC(ft_richAdd,ft_richSmooth)
```

It looks like the model with additive terms is a better fit.

*We could try refitting the model to account for this, using the `gamm` function in the `mgcv` package. Do you think this would make any difference to our conclusion? Justify your argument.*

I'm not going to try refitting, I don't think it would make any difference. The main thing that happens if you have spatial structure in your data that you haven't accounted for is that you get *false confidence*: standard errors and *P*-values are too small. In a model selection context, this means that you can end up saying terms should be in the model that you don't actually need in there, so you end up with an overly complicated  model. But in this case, we have already concluded that we don't need the more complicated model, with the extra smoother terms in it.



## Exercise 8.5: Carbon dioxide measurements at Mauna Loa observatory
*We would like to model the dataset to characterise the key trends within as well as across years, taking into account that carbon dioxide does not follow a linear trend over time, and that there is a periodic seasonal oscillation... What sort of model would you use?*

We can fit an additive model with a smoother for the long-term trend, and harmonics (sin/cos terms) for the seasonal trend.


## Code Box 8.6: A simple model for the Mauna Loa monthly data with a cyclical predictor
```{r box86, fig.width=7}
data(maunaloa)
library(mgcv)
ft_cyclic=gam(co2~s(DateNum)+sin(month/12*2*pi)+cos(month/12*2*pi),
  data=maunaloa)
plot(maunaloa$co2~maunaloa$Date,type="l",
  ylab=expression(CO[2]),xlab="Time")
points(predict(ft_cyclic)~maunaloa$Date,type="l",col="red",lwd=0.5)
```


## Code Box 8.7: Residual plots across time and season for the Mauna Loa monthly data

```{r box87, fig.width=9}
par(mfrow=c(1,2))
plot(residuals(ft_cyclic)~maunaloa$Date,type="l", xlab="Time")
plot(residuals(ft_cyclic)~sin(maunaloa$month/12*2*pi),
       type="l",xlab="Season")
```


Here is some code to produce a simulated dataset from this model, for comparison:
```{r box87 sim, fig.width=9}
maunaloa$simCO2 = unlist(simulate(ft_cyclic))
ft_simCyclic=gam(simCO2~s(DateNum)+sin(month/12*2*pi)+cos(month/12*2*pi),
  data=maunaloa)
par(mfrow=c(1,2))
plot(residuals(ft_simCyclic)~maunaloa$Date,type="l", xlab="Time")
plot(residuals(ft_simCyclic)~sin(maunaloa$month/12*2*pi),
       type="l",xlab="Season")
```


## Code Box 8.8: Another model for the Mauna Loa monthly data, with an extra sine curve in there to better handle irregularities in the seasonal effect}\label{code:maunaAnotherModel}

```{r box88, fig.width=9}
ft_cyclic2=gam(co2~s(DateNum)+sin(month/12*2*pi)+cos(month/12*2*pi)+
       sin(month/12*4*pi)+cos(month/12*4*pi),data=maunaloa)
par(mfrow=c(1,2))
plot(residuals(ft_cyclic2)~maunaloa$Date,type="l", xlab="Time")
plot(residuals(ft_cyclic2)~sin(maunaloa$month/12*2*pi),
       type="l",xlab="Season")
```


## Code Box 8.9: Mauna Loa model with autocorrelation}\label{code:maunaAutoModel}

```{r box89}
ft_gamm = gamm(co2~s(DateNum)+sin(month/12*2*pi)+cos(month/12*2*pi)+
                sin(month/12*4*pi)+cos(month/12*4*pi),correlation=corAR1(),
                data=maunaloa)
acf(residuals(ft_gamm$gam))
acf(residuals(ft_gamm$lme,type="normalized"))
```

## Exercise 8.6: Mauna Loa monthly data -- an extra term in seasonal trend?
*We tried modelling the trend by fitting sine curves with two frequencies... Is this sufficient, or should we add another frequency too, like `month/12*6*pi` (for curves with three cycles a year)?*

```{r ex86, fig.width=9}
ft_cyclic3=gam(co2~s(DateNum)+sin(month/12*2*pi)+cos(month/12*2*pi)+
       sin(month/12*4*pi)+cos(month/12*4*pi) +
       + sin(month/12*6*pi)+cos(month/12*6*pi), data=maunaloa)
ft_cyclic4=gam(co2~s(DateNum)+sin(month/12*2*pi)+cos(month/12*2*pi)+
       sin(month/12*4*pi)+cos(month/12*4*pi) +
       + sin(month/12*6*pi)+cos(month/12*6*pi)+
         sin(month/12*8*pi)+cos(month/12*8*pi), data=maunaloa)
par(mfrow=c(1,2))
plot(residuals(ft_cyclic4)~maunaloa$Date,type="l", xlab="Time")
plot(residuals(ft_cyclic4)~sin(maunaloa$month/12*2*pi),
       type="l",xlab="Season")
BIC(ft_cyclic,ft_cyclic2,ft_cyclic3,ft_cyclic4)
```

According to BIC, 2 harmonics is the winner, there isn't much to gain from adding a third or fourth.


## Exercise 8.7: Mauna Loa annual data -- temporal autocorrelation?}
*Consider the annual Mauna Loa monthly data of Code Box 8.1, which uses the measurements from January only... Is there evidence of temporal autocorrelation in this dataset?*

```{r ex87}
maunaJan = maunaloa[maunaloa$month==1,]
ft_gammJan = gamm(co2~s(year),correlation=corAR1(), data=maunaJan)
acf(residuals(ft_gammJan$gam))
acf(residuals(ft_gammJan$lme,type="normalized"))
ft_gammJan$lme
```

There isn't really an autocorrelation signal here -- the lags represent years now rather than months, and consecutive years (lag 1) have a correlation of less than 0.1 so it looks like not much is happening. The mixed effects model doesn't really do much here. The estimated autocorrelation is surprisingly large given this (0.26) but it is much smaller than in the monthly analysis (0.76) and decays to negligible levels quickly (correlation at two years is 0.26^2=0.07).


## Exercise 8.8: Aspect as a predictor of species richness
*Consider Ian's species richness data... How would you enter `aspect` into the model?*
This is a cyclical variable so we would add `cos` and `sin` terms to the model.

```{r ex88 aspect}
data(Myrtaceae)
summary(Myrtaceae$aspect)
```

It seems to have been entered in degrees so has period 360. Maybe we can consider two harmonics to try to account for non-sinusoidal response. We will keep soil, temperature and rainfall variables from previously:

```{r ex88 period}
data(Myrtaceae)
Myrtaceae$logrich=log(Myrtaceae$richness+1)
ft_richAsp = lm(logrich~soil+poly(TMP_MAX,degree=2)+
     poly(TMP_MIN,degree=2)+poly(RAIN_ANN,degree=2)+
               cos(aspect*2*pi/360)+sin(aspect*2*pi/360)+cos(aspect*4*pi/360)+sin(aspect*4*pi/360),
     data=Myrtaceae)
ft_richAdd = lm(logrich~soil+poly(TMP_MAX,degree=2)+
     poly(TMP_MIN,degree=2)+poly(RAIN_ANN,degree=2), data=Myrtaceae)
BIC(ft_richAsp,ft_richAdd)
```

It looks like there isn't really an `aspect` signal in the data, BIC is much smaller without any aspect terms being in the model.

*Recall that species richness had some spatial structure to it. Refit the model using the `gls` function from the `nlme` package to account for spatially structured correlation. Did this change your inferences about `aspect` in any appreciable way?*

Well it won't change our inferences because ignoring spatial correlation risks giving you false confidence in the importance of terms, but we already decided that `aspect` is not important. But anyway. We decided last time around that an exponential model with a nugget effect did OK:

```{r ex88 spatial}
library(nlme)
ft_richNugg = gls(logrich~soil+poly(TMP_MAX,degree=2)+poly(TMP_MIN,degree=2)+
       poly(RAIN_ANN,degree=2),data=Myrtaceae,
         correlation=corExp(form=~X+Y,nugget=TRUE))
ft_richAspNugg = gls(logrich~soil+poly(TMP_MAX,degree=2)+poly(TMP_MIN,degree=2)+
       poly(RAIN_ANN,degree=2)+ cos(aspect*2*pi/360)+sin(aspect*2*pi/360)+cos(aspect*4*pi/360)+sin(aspect*4*pi/360),data=Myrtaceae,
         correlation=corExp(form=~X+Y,nugget=TRUE))
BIC(ft_richNugg)
BIC(ft_richAspNugg)
```

As expected, this did not change inference in any appreciable way!
`

