---
title: "Eco-Stats - Code and Data Accompanying the Eco-Stats Text"
author: "David Warton"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Eco-Stats - Code and Data Accompanying the Eco-Stats Text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The Eco-Stats package contains functions and data supporting the Eco-Stats text (Warton, 2022, Springer), and solutions to exercises. Functions include tools for using simulation envelopes in diagnostic plots, also applicable to multivariate linear models, and a parametric bootstrap function that can be used in place of an `anova` call for hypothesis testing via simulation, for many types of regression models. Datasets mentioned in the package are included here (where not available elsewhere) and vignettes work through code chunks and exercises from the textbook, one chapter at a time.

## Simulation envelopes in plots

The command `plotenvelope` will take standard residual plots for most model objects, but add global envelopes around points (for quantile plots) and around smoothers (for residual plots) constructed via simulation. These are constructed using the `GET` package as global envelopes, that is, when model assumptions are correct, 95% of quantile plot envelopes (at confidence level 95%) should contain _all_ datapoints.

```{r, echo=TRUE, results='asis'}
library(ecostats)
data(iris)
Y = with(iris, cbind(Sepal.Length,Sepal.Width,Petal.Length,Petal.Width))
iris.mlm=lm(Y~Species,data=iris)
# check normality assumption:
plotenvelope(iris.mlm,n.sim=199)
```
For `mlm` objects, this function will compute conditional residuals and fitted values, that is, they are computed for each response conditional on all other responses being observed, via the `cpredict` and `cresiduals` functions. This is done because the full conditionals of a distribution are known to be diagnostic of joint distributions, hence any violation of the multivariate normality assumption will be expressed as a violation of assumptions of these full conditional models. The full conditionals are well-known to follow a linear model, as a function of all other responses as well as predictors.

The `qqenvelope` function can be applied for a normal quantile plot, with global envelope, to either a fitted model or a sample of data:
```{r, echo=TRUE, results='asis'}
y=rnorm(20)
qqenvelope(y)
```


## Datasets

All datasets used in the Eco-Stats text, where not available elsewhere, are supplied here:
```{r, echo=TRUE, results='asis'}
 data(aphids)
 cols=c(rgb(1,0,0,alpha=0.5),rgb(0,0,1,alpha=0.5)) #transparent colours
 with(aphids$oat, interaction.plot(Time,Plot,logcount,legend=FALSE,
                                col=cols[Treatment], lty=1, ylab="Counts [log(y+1) scale]",
                                xlab="Time (days since treatment)") )
 legend("bottomleft",c("Excluded","Present"),col=cols,lty=1)
```
